Curve counting in genus one: \\ elliptic singularities {\it \&} relative geometry

### 2020-10-16 EllipticSingularitiesSplitting.tex
minor changes to S1, in particular added a different formulation of pf(thm 1.4)

\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{faktor}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{tikz,tikz-cd}
\usetikzlibrary{decorations.pathmorphing,decorations.pathreplacing,patterns,arrows.meta}

@@

The higher genus situation is much more delicate, and substantial theory has been developed in recent years to treat the genus one case, both for its intrinsic geometry, and for higher genus insights. This began over a decade ago with pioneering work of Li, Vakil, and Zinger~\cite{redgone,VZ,LZ,lz2,zingerstvsred,zingred}, and was recently reinvigorated using ideas from singularity theory~\cite{BCM18,HL,SMY1,VISC} and logarithmic geometry~\cite{RSPW,RSPW2}. The result is a \textbf{reduced} Gromov--Witten theory, which removes degenerate contributions from the ordinary theory. While the Atiyah--Bott localisation for the reduced invariants was used to great effect by Zinger in his proof of the BCOV conjectures, two major components of the subject -- relative invariants and the degeneration formula -- have not been developed in this setting. Our work provides these as new {structural} tools in reduced genus one Gromov--Witten theory\footnote{Further motivation to consider the genus one case specifically comes simply from the fact that for Calabi--Yau geometries in dimension larger than $3$, Gromov--Witten theory vanishes in genera greater than one.}. While the essential starting point is the early work on relative Gromov--Witten theory by Gathmann, modern tools are required in our setting. We establish both the conceptual framework and the formula by combining tropical geometry and logarithmic Gromov--Witten theory, as developed by Abramovich--Chen--Gross--Siebert, with the recent understanding of elliptic singularities and their interactions with logarithmic structures.

@@

\begin{customthm}{A}\label{thm: desingularization}
There exists a logarithmically smooth and proper Deligne--Mumford stack $\VZ_{1,\alpha}(\mathbb P^N|H,d)$ having expected dimension, and containing $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$  as a dense open subset. A point in this moduli space determines commutative diagrams
\[
\begin{tikzcd}
C\arrow{r} & \overline C_1\arrow{d}\arrow{r} & \overline C_2\arrow{d} \\ 
& \mathbb P^N[s]\arrow{r} & \mathbb P^N,
\end{tikzcd}
\]
where $C$ is a nodal genus $1$ curve, $\overline C_1$ and $\overline C_2$ are Gorenstein curves of genus $1$, the map $C\to \PP^N[s]$ is a logarithmic map to an $s$-fold expansion of $\PP^N$ along $H$, and each of the vertical is \textbf{non-degenerate}.
\end{customthm}

Intuitively, non-degeneracy is the condition that the maps in question are non-constant on some branch of the minimal genus $1$ subcurve, see  \S \ref{section construction} for details.

@@

\subsection{Context and further discussion} This paper is a most directly related to recent work on aligned logarithmic structures, elliptic singularities, and stable maps, as developed in~\cite{RSPW,RSPW2}. The moduli spaces we construct lie in between the absolute and toric-boundary-relative geometries studied in the first and second of those papers, respectively. The degeneration algorithm presented here as a first practical implementation of the theory developed in that work.

@@

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Pierrick Bousseau, Francesca Carocci, Cristina Manolache, Davesh Maulik, Keli Santos-Parker, and Jonathan Wise. The comments of an anonymous referee have proved very valuable. The authors worked on this project during visits to MPIM-Bonn, MIT, University of Cambridge, and the University of Glasgow, and thank these institutions for ideal working conditions and support. L.B. and N.N. would in addition like to thank Imperial College London, the LSGNT, the Royal Society, 3CinG, and the MSRI program ''Enumerative Geometry beyond Numbers''. N.N. was partially supported by EPSRC grant EP/R009325/1.

@@

\subsection*{Curves and tropical curves} A nodal curve $C$ over $\spec \mathbb C$ has an associated dual graph $\Gamma(C)$. Given a toric monoid $P$, there is a logarithmic point $\spec \CC_P$ which may be thought of as remembering the information of an embedding of a closed point into the toric variety $\spec \mathbb C[P]$. For a logarithmically smooth curve $C$ over $\spec \CC_P$, the dual graph $\Gamma(C)$ is enhanced: given an edge $e$ of $\Gamma(C)$, the logarithmic structure keeps track of a generalized ``edge length'' $\ell_e$, which is an element of $P$. These data can be repackaged into a cone complex $\plC$ together with a map to the dual cone $\sigma_P$
\[
\pi: \plC\to \sigma_P.
\]
A fiber of this map $\pi$ is a metric space enhancing the topological space $\Gamma(C)$, with edges given lengths in $\RR_{\geq 0}$. Over the interior of the cone $\sigma_P$ these edge lengths are nonzero, and the fibers are homoemorphic to $\Gamma(C)$. Over the faces of $\sigma_P$, the fibers can be identified with edge collapses.

The discussion is globalized in~\cite{CavalieriChanUlirschWise}, and the construction can be carried out for the universal curve over the moduli space of logarithmic curves. Specifically, there is a stack over the category of cone complexes $\mathfrak M_{g,n}^{\mathrm{trop}}$ equipped with a universal curve 
\[
\plC\to \mathfrak M_{g,n}^{\mathrm{trop}}.
\]
The reader is encourage to think of $\mathfrak M_{g,n}^{\mathrm{trop}}$ as playing the same role for $\mathfrak M_{g,n}$ as a fan plays for a toric variety. In particular, two constructions are relevant:
\begin{enumerate}
\item A subdivision of $\mathfrak M_{g,n}^{\mathrm{trop}}$ gives rise to a birational modification of $\mathfrak M_{g,n}$.
\item A piecewise linear function on $\mathfrak M_{g,n}^{\mathrm{trop}}$ gives rise to a Cartier divisor on $\mathfrak M_{g,n}$.
\end{enumerate}

\subsection*{Maps and tropical maps}The discussion carries over to the setting of logarithmic maps. Specifically, given a logarithmic curve $C$ and a logarithmic map to a smooth pair $(X,Y)$ over a logarithmic scheme $S$. For simplicity, assume that $S$ is a $P$-logarithmic point, as above. There is an associated \textbf{tropicalization}, i.e. a family of tropical curves $\plC$ over $\sigma_P$, together with a map of cone complexes
\[
\plC\to \RR_{\geq 0}.
\]
Each fiber over $\sigma_P$ encodes a metric enhancement of the dual graph, together with a piecewise linear map to $\RR_{\geq 0}$. There is once again a cone stack representing the moduli problem of tropical maps; the technical details are only cosmetically different than those presented in~\cite{CavalieriChanUlirschWise}, and we leave these details to the reader.

\subsection{Maps from curves and elliptic singularities} We provide a brief intuitive picture that underlies the work in~\cite{RSPW,RSPW2}. If $C$ is a Gorenstein curve and $f:C\to \mathbb P^N$ is a morphism, the ampleness of the the tangent bundle of $\mathbb P^N$ ensures that if no subcurve of arithmetic genus $1$ is contracted, the deformations of $f$ are unobstructed. In fact, the deformations \textit{over the stack of curves} are also unobstructed. The moduli space of Gorenstein genus $1$ is not smooth. However, a more sophisticated moduli problem, consisting of nodal curves together with a \textit{chosen contraction} to a Gorenstein curve, might be expected to resolve these singularities. The formation of such a moduli problem requires logarithmic geometry, and is achieved in~\cite[Section~3]{RSPW}. Once constructed, the corresponding moduli of stable maps out of a Gorenstein curve is also resolved. Note that maps from nodal genus $1$ curves may have stable limits that contract genus $1$ subcurves, but the allowance of Gorenstein singularities ensures that properness despite disallowing contracted genus $1$ components.

When working with logarithmic geometries such as $(\mathbb P^N,H)$ as we do here, similar ideas may be expected to apply, but the deformation theory of maps becomes more delicate. Indeed, the logarithmic Euler sequence for this pair shows that the logarithmic tangent bundle is a quotient of $\mathcal O\oplus\mathcal O(1)^{\oplus 1}$, and additional obstructions appear. In simple terms, there exist logarithmic maps to the pair $(\mathbb P^N,H)$ with obstructed logarithmic deformations, whose underlying ordinary stable map is unobstructed.

@@

Consider a logarithmically smooth curve $C$ over a logarithmic geometric point $S$ with monoid $P$ and let $\pi: \plC\to \sigma_P$ denote its tropicalization. Given a point $t$ in $\sigma_P$ we obtain a metric graph as explained above. Since the curve in question has genus $1$, there is a minimal genus $1$ subgraph -- either a vertex of genus $1$ or a cycle in the graph. We will refer to this as the \textbf{core}. Given a point $s$ with $\pi(s)$ equal to $t$, there is a well-defined distance in the metric graph $\pi^{-1}(t)$ from $s$ to the core.

In particular, for each vertex $v$ of the dual graph $\Gamma(C)$, i.e. each component of $C$, there is a well-defined function $\lambda(v): \sigma_P\to \RR_{\geq 0}$, defined on the base of the tropical family. Given any other piecewise linear function $\delta$ on $\sigma_P$, say that $\delta$ is \textbf{comparable} to $\lambda(v)$ if there is a global order relation, either $\lambda(v)\leq \delta$ or $\lambda(v)\geq \delta$, that holds at all points of $\sigma_P$.

Piecewise linear functions on $\sigma_P$ are naturally identified with global sections of the characteristic sheaf $\overline M_S$. Given an arbitrary logarithmically smooth curve $C/S$, conditions involving the comparability of piecewise linear functions can be imposed geometric point-by-point. This leads to the following definition.

@@

The moduli stack $\mathfrak M_{1,n}^{\mathrm{cen}}$  of centrally aligned curves is a logarithmic algebraic stack in the smooth topology, and $\mathfrak M_{1,n}^{\mathrm{cen}}\to\mathfrak M_{1,n}$ is a logarithmic modification. The main construction of~\cite{RSPW} canonically associates to any centrally aligned family of curves $\mathcal C_S$ a partial destabilisation $\widetilde{\mathcal C}_S$, introducing $2$-pointed components, and a contraction
\[
\widetilde{\mathcal C}_S\to \overline{\mathcal C}_S,
\]
where $\overline{\mathcal C}_S$ may contain a Gorenstein elliptic singularity \cite{SMY1}. Given a family of tropical curves, the function $\delta$ provides a non-negative real number for each point on the base. We may therefore consider the ``circle'' of points on each fiber whose distance from the minimal genus $1$ subgraph is equal to $\delta$. The partial destabilization, or ``bubbling'', introduces $2$-pointed rational curves into the curve by adding vertices to the dual graph at the points that lie on this circle. The number of branches is equal to the number of vertices of $\widetilde{\sqC}$ which lie on the circle of radius $\delta$. %This data uniquely determines the singularity.

The space of stable maps $\widetilde\VZ_{1,n}(\PP^N,d)$ from the universal nodal curve $\mathcal C$ over the stack $\mathfrak M_{1,n}^{\mathrm{cen}}$ is proper and algebraic, with projective coarse moduli space. A compatibility condition between the alignment and the map is required, namely that the circle of radius $\delta$ passes through at least one component of positive degree. A stable map in this space is said to satisfy the \textbf{factorisation condition} if the map $\mathcal C\to \mathbb P^N$ factors through the contraction $\widetilde{\mathcal C}\to \overline{\mathcal C}$. For ease of language, maps from nodal curves that are not constant on any subcurve of genus $1$ are taken to satisfy this condition.

@@

\begin{remark}[Constant maps]\label{rem: constant-maps}
When the degree of the map is $1$, maps from smooth genus $1$ curves to $\PP^N$ are constant. The interior of the moduli space of stable maps in this case is empty, though the compactification by stable maps is non-empty. The space of maps satisfying the factorization condition is also empty. Only non-empty cases are considered. 
\end{remark}

@@

\subsection{Relative geometry: compactification} Fix a hyperplane $H\subseteq \mathbb P^N$. Let $\alpha$ be a partition of the degree $d>0$. Consider the moduli space $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$ of maps from smooth elliptic curves $C\to \mathbb P^N$ that meet $H$ at finitely many marked points with vanishing orders given by the partition $\alpha$. This is a smooth, but non-proper Deligne--Mumford stack.

@@

In order make the obstruction theory more geometric, we expand the target. To elucidate the connection with the static target, consider a logarithmic stable map $[C\to (\mathbb P^N,H)]$ over $\mathrm{Spec} \ (\mathbb N\to \mathbb C)$. At the level of tropicalizations, there is a map of fans
\[
\plC\to \mathbb R_{\geq 0}.
\]
Choose a subdivision of $\mathbb R_{\geq 0}$ whose vertices are the images of vertices of $\plC$. Pull this subdivision back to $\plC$ by subdividing preimages of the vertices of $\mathbb R_{\geq 0}$. Denote the resulting map $\widetilde \plC \to \widetilde{\mathbb R}_{\geq 0}$. These subdivisions induce logarithmic modifications
\[
\widetilde C\to \mathbb P^N[s],
\]
see~\cite{AW}. Here the latter is the $s$-times iterated deformation to the normal cone of $H$ in $\mathbb P^N$. The irreducible components of the target are in bijection with the vertices in $\widetilde{\mathbb R}_{\geq 0}$. The curve is modified by adding rational components for the newly introduced vertices.

@@

Globally, Kim constructs a moduli space of logarithmic maps to expanded degenerations, which on logarithmic points, reduces to the above construction~\cite{KimLog}. Kim's space is identified with a subcategory of the Abramovich--Chen--Gross--Siebert space, and its dual minimal monoids are cones in a subdivision of the dual minimal cones of the unexpanded space, see~\cite[\S~2]{R19}.

@@

\subsection{Relative geometry: factorisation}\label{subsection factorisation} Let $[C\to \mathbb P^N[s]\to \mathbb P^N]$ be a centrally aligned logarithmic map. Recall that $\mathbb P^N[s]$ consists of a union of $\mathbb P^N$ with $s$ copies of the projective bundle $\mathbb{P} = \mathbb P(\mathcal O_H\oplus \mathcal O_H(1))$. We refer to these latter components as \textbf{the higher levels}. Thus, we will say that a subcurve $D\subseteq C$ maps to \textbf{higher level} if the composite $C\to \mathbb P^N$ maps $D$ into $H\subset \mathbb P^N$.

Let $D_F\subseteq C$ be the maximal genus one subcurve which is mapped to a higher level and contracted by the map $C\to \mathbb P^N[s]$. Let $\delta_F$ denote the distance from the core to the nearest component of $C \setminus D_F$. Similarly let $D_B\subseteq C$ be the maximal genus one subcurve that is contracted by the collapsed map to $\mathbb P^N$ and let $\delta_B$ be the associated radius. This coincides with the radius of the underlying map to $\mathbb P^N$. Of course, $\delta_F\leq \delta_B$, so $(C,\delta_F)$ is centrally aligned.

@@

A Gorenstein elliptic singularity has a number of \textbf{branches}. Given a curve $\overline C$ of genus $1$ with such a singularity, define a branch of $\overline C$ to be an irreducible component containing a branch of the singularity. If the curve $C$ is nodal, then its branches will be understood to be the components of its minimal arithmetic genus $1$ subcurve, i.e. its core. The following condition identifies the locus of smoothable maps.

\begin{definition}
The map $[C\to \mathbb P^N[s]\to \mathbb P^N]$ \textbf{factorises completely} if 
\begin{enumerate}
\item the map $C\to \mathbb P^N[s]$ factors through $\overline C_F$ such that $F$ is non-constant on at least one branch of $\overline C_F$,
\item the collapsed map to $\mathbb P^N$ factorises through $\overline C_B$ such that at least one branch of $\overline C_B$ has positive degree.
\end{enumerate}
The factorisation conditions are understood to be satisfied when $C\to \mathbb P^N[s]$ (resp. $C\to \mathbb P^N$) does not contract a genus $1$ subcurve. 
\end{definition}

>>>>

\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{faktor}
\usepackage{xcolor}
\usepackage{xfrac}
\usepackage{tikz,tikz-cd}
\usetikzlibrary{decorations.pathmorphing,decorations.pathreplacing,patterns,arrows.meta,shapes.misc}

@@

The higher genus situation is much more delicate, and substantial theory has been developed in recent years to treat the genus one case, both for its intrinsic geometry, and for higher genus insights. This began over a decade ago with pioneering work of Li, Vakil, and Zinger~\cite{redgone,VZ,LZ,lz2,zingerstvsred,zingred}, and was recently reinvigorated using ideas from singularity theory~\cite{BCM18,HL,SMY1,VISC} and logarithmic geometry~\cite{RSPW,RSPW2,BC20}. The result is a \textbf{reduced} Gromov--Witten theory, which removes degenerate contributions from the ordinary theory. While the Atiyah--Bott localisation for the reduced invariants was used to great effect by Zinger in his proof of the BCOV conjectures, two major components of the subject -- relative invariants and the degeneration formula -- have not been developed in this setting. Our work provides these as new {structural} tools in reduced genus one Gromov--Witten theory\footnote{Further motivation to consider the genus one case specifically comes simply from the fact that for Calabi--Yau geometries in dimension larger than $3$, Gromov--Witten theory vanishes in genera greater than one.}. While the essential starting point is the early work on relative Gromov--Witten theory by Gathmann, modern tools are required in our setting. We establish both the conceptual framework and the formula by combining tropical geometry and logarithmic Gromov--Witten theory, as developed by Abramovich--Chen--Gross--Siebert, with the recent understanding of elliptic singularities and their interactions with logarithmic structures.

@@

\begin{customthm}{A}\label{thm: desingularization}
There exists a logarithmically smooth and proper Deligne--Mumford stack $\VZ_{1,\alpha}(\mathbb P^N|H,d)$ having expected dimension, and containing $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$  as a dense open subset. A point in this moduli space determines commutative diagrams
\[
\begin{tikzcd}
C\arrow{r} & \overline C_1\arrow{d}\arrow{r} & \overline C_2\arrow{d} \\ 
& \mathbb P^N[s]\arrow{r} & \mathbb P^N,
\end{tikzcd}
\]
where $C$ is a nodal genus $1$ curve, $\overline C_1$ and $\overline C_2$ are Gorenstein curves of genus $1$, the map $C\to \PP^N[s]$ is a logarithmic map to an $s$-fold expansion of $\PP^N$ along $H$, and each of the vertical maps does not contract the minimal subcurve of genus $1$.
\end{customthm}

%Intuitively, non-degeneracy is the condition that the maps in question are non-constant on some branch of the minimal genus $1$ subcurve, see  \S \ref{section construction} for %details.

@@

\subsection{Context and further discussion} This paper is most directly related to recent work on aligned logarithmic structures, elliptic singularities, and stable maps, as developed in~\cite{RSPW,RSPW2}. The moduli spaces we construct lie in between the absolute and toric-boundary-relative geometries studied in the first and second of those papers, respectively. The degeneration algorithm presented here as a first practical implementation of the theory developed in that work.

@@

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Pierrick Bousseau, Francesca Carocci, Cristina Manolache, Davesh Maulik, Keli Santos-Parker, and Jonathan Wise. The comments of an anonymous referee have proved very valuable. The authors worked on this project during visits to MPIM-Bonn, MIT, University of Cambridge, and the University of Glasgow, and thank these institutions for ideal working conditions and support. L.B. and N.N. would in addition like to thank Imperial College London, the LSGNT, the Royal Society, 3CinG, and the MSRI program ''Enumerative Geometry beyond Numbers''. N.N. was partially supported by EPSRC grant EP/R009325/1.  L.B. is supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Germany’s Excellence Strategy EXC-2181/1 - 390900948 (the Heidelberg STRUCTURES Cluster of Excellence).

@@

\subsection*{Curves and tropical curves} A nodal curve $C$ over $\spec \mathbb C$ has an associated dual graph $\Gamma(C)$. Given a toric monoid $P$, there is a logarithmic point $\spec \CC_P$ which may be thought of as remembering the information of the embedding of the torus-fixed closed point into the toric variety $\spec \mathbb C[P]$. For a logarithmically smooth curve $C$ over $\spec \CC_P$, the dual graph $\Gamma(C)$ is enhanced: given an edge $e$ of $\Gamma(C)$, the logarithmic structure keeps track of a generalized ``edge length'' $\ell_e$, which is an element of $P$. These data can be repackaged into a cone complex $\plC$ together with a map to the dual cone $\sigma_P$
\[
\pi: \plC\to \sigma_P.
\]
A fiber of this map $\pi$ is a metric space enhancing the topological space $\Gamma(C)$, with edges given lengths in $\RR_{\geq 0}$. Over the interior of the cone $\sigma_P$ these edge lengths are nonzero, and the fibers are homoemorphic to $\Gamma(C)$. Over the faces of $\sigma_P$, the fibers can be identified with edge contractions.

The discussion is globalized in~\cite{CavalieriChanUlirschWise}, and the construction can be carried out for the universal curve over the moduli space of logarithmic curves. Specifically, there is a stack over the category of cone complexes $\mathfrak M_{g,n}^{\mathrm{trop}}$ equipped with a universal curve 
\[
\plC\to \mathfrak M_{g,n}^{\mathrm{trop}}.
\]
The reader is encouraged to think of $\mathfrak M_{g,n}^{\mathrm{trop}}$ as playing the same role for $\mathfrak M_{g,n}$ as a fan plays for a toric variety. In particular, two constructions are relevant:
\begin{enumerate}
\item A subdivision of $\mathfrak M_{g,n}^{\mathrm{trop}}$ gives rise to a birational modification of $\mathfrak M_{g,n}$.
\item A piecewise linear function on $\mathfrak M_{g,n}^{\mathrm{trop}}$ gives rise to a Cartier divisor on $\mathfrak M_{g,n}$.
\end{enumerate}

\subsection*{Maps and tropical maps}The discussion carries over to the setting of logarithmic maps. Specifically, given a logarithmic curve $C$ and a logarithmic map to a smooth pair $(X,Y)$ over a logarithmic scheme $S$. For simplicity, assume that $S$ is a $P$-logarithmic point, as above. There is an associated \textbf{tropicalization}, i.e. a family of tropical curves $\plC$ over $\sigma_P$, together with a map of cone complexes
\[
\plC\to \RR_{\geq 0}.
\]
Each fiber over $\sigma_P$ encodes a metric enhancement of the dual graph, together with a piecewise linear map to $\RR_{\geq 0}$. There is once again a cone stack representing the moduli problem of tropical maps; the technical details are only cosmetically different than those presented in~\cite{CavalieriChanUlirschWise}, and we leave these details to the reader. See also \cite{GrossSiebertLog}.

\subsection{Maps from curves and elliptic singularities} We provide a brief intuitive picture that underlies the work in~\cite{RSPW,RSPW2}. If $C$ is a Gorenstein curve and $f:C\to \mathbb P^N$ is a morphism, the ampleness of the the tangent bundle of $\mathbb P^N$ ensures that if no subcurve of arithmetic genus $1$ is contracted, the deformations of $f$ are unobstructed. In fact, the deformations \textit{over the stack of curves} are also unobstructed. The moduli space of Gorenstein genus $1$ curves is not smooth. However, a more sophisticated moduli problem, consisting of nodal curves together with a \textit{chosen contraction} to a Gorenstein curve, might be expected to resolve these singularities. The formation of such a moduli problem requires logarithmic geometry, and is achieved in~\cite[Section~3]{RSPW}. Once constructed, the corresponding moduli of stable maps out of a Gorenstein curve is also resolved. Note that maps from nodal genus $1$ curves may have stable limits that contract genus $1$ subcurves, but the allowance of Gorenstein singularities ensures properness despite disallowing contracted genus $1$ components.

When working with logarithmic geometries such as $(\mathbb P^N,H)$, as we do here, similar ideas may be expected to apply, but the deformation theory of maps becomes more delicate. Indeed, the logarithmic Euler sequence for this pair shows that the logarithmic tangent bundle is a quotient of $\mathcal O\oplus\mathcal O(1)^{\oplus N}$, and additional obstructions appear. In simple terms, there exist logarithmic maps to the pair $(\mathbb P^N,H)$ with obstructed logarithmic deformations, whose underlying ordinary stable map is unobstructed.

@@

Consider a logarithmically smooth curve $C$ over a logarithmic geometric point $S$ with monoid $P$ and let $\pi\colon \plC\to \sigma_P$ denote its tropicalization. Given a point $t$ in $\sigma_P$ we obtain a metric graph as explained above. Since the curve in question has genus $1$, there is a minimal genus $1$ subgraph -- either a vertex of genus $1$ or a cycle in the graph. We will refer to this as the \textbf{core}.

Given a point $s$ with $\pi(s)$ equal to $t$, there is a well-defined distance in the metric graph $\pi^{-1}(t)$ from $s$ to the core. 
In particular, for each vertex $v$ of the dual graph $\Gamma(C)$, i.e. each component of $C$, there is a well-defined function $\lambda(v): \sigma_P\to \RR_{\geq 0}$, defined on the base of the tropical family. We may as well think of $\boldsymbol{\lambda}$ as a piecewise linear function $\lambda\colon\plC\to\RR_{\geq 0}\times\sigma_P$, see Figure \ref{fig:lambdaexample}. Given any other linear function $\delta$ on $\sigma_P$, say that $\delta$ is \textbf{comparable} to $\lambda(v)$ if there is a global order relation, either $\lambda(v)\leq \delta$ or $\lambda(v)\geq \delta$, that holds at all points of $\sigma_P$.

Piecewise linear functions on $\plC$ are naturally identified with global sections of the characteristic sheaf $\overline M_C$; linear functions on $\sigma_P$ are naturally identified with global sections of the characteristic sheaf $\overline M_S$, and can be pulled back to $C$. Given an arbitrary logarithmically smooth curve $C/S$, conditions involving the comparability of piecewise linear functions can be imposed at the level of geometric points; they determine subdivisions of $\sigma_P$, and, in turn, logarithmic modifications of $S$. %This leads to the following definition.

@@

\begin{figure}
 \begin{tikzpicture} 
  \tikzset{cross/.style={cross out, draw=black, fill=none, minimum size=2*(#1-\pgflinewidth), inner sep=0pt, outer sep=0pt}, cross/.default={3pt}}

  \foreach \x in {(.5,2.5),(3.3,2.5),(3.65,2.5),(3.7,2.5),(4.4,2.5),(4.65,2.5),(-1.7,2.5)}
  \draw \x node[cross,rotate=45] {};

  \foreach \x in {(-1,1),(3.5,1.5),(3.2,2),(-2.5,2.5),(-.5,1.5),(1.5,3.5),(4.8,3),(5.8,3.5),(7,0),(-2,2.8),(4,3.2)}
  \draw[fill=black] \x circle (2pt);

  \draw[gray] (0,0) node[left]{$\plC_0$} -- (1,.2) -- (2,0) -- (1,-.2) -- (0,0); %core

  \draw (0,0) -- (-2.5,2.5) -- (-2.7,4) (-2.5,2.5) -- (-2.3,4) (-1,1)--(1.5,3.5)--(1.3,4) (1.5,3.5)--(1.7,4) (-.5,1.5) -- (-1.7,2.5) -- (-2,2.8) --(-2.2,4) (-2,2.8) --(-1.8,4);
  \draw (1,.2) -- (4,3.2) -- (3.9,4) (4,3.2) -- (4.1,4);
  \draw (1,-.2) -- (3.2,2) -- (5,4) (3.2,2) -- (5.2,4);
  \draw (2,0) -- (3.5,1.5) -- (4.8,3) -- (5.4,4) (4.8,3) -- (5.6,4) (3.5,1.5) -- (5.8,3.5) -- (5.8,4) (5.8,3.5) -- (6,4);

  \draw[->] (7,0) -- (7,4); %real axis
  \draw[dashed] (-2.7,1) -- (7.2,1) (-2.7,1.5) -- (7.2,1.5) (-2.7,2) -- (7.2,2); %heights
  \draw[blue,dashed]  (-2.7,2.5) -- (7.2,2.5) node[right]{$\delta$};
  \draw [->] (6,1.55) --node[above]{$\lambda$} (6.8,1.55); %arrow-lambda

  \foreach \x in {(0,0),(1,.2),(1,-.2),(2,0)}
  \draw[fill=gray] \x circle (2pt);
 \end{tikzpicture}
\caption{The function $\lambda$ (distance from the core) and its cutoff value $\delta$.}
\label{fig:lambdaexample}
\end{figure}

The moduli stack $\mathfrak M_{1,n}^{\mathrm{cen}}$  of centrally aligned curves is a logarithmic algebraic stack in the smooth topology, and $\mathfrak M_{1,n}^{\mathrm{cen}}\to\mathfrak M_{1,n}$ is a logarithmic modification. The main construction of~\cite{RSPW} canonically associates to any centrally aligned family of curves $\mathcal C_S$ a partial destabilisation $\widetilde{\mathcal C}_S$, introducing $2$-pointed components, and a contraction
\[
\widetilde{\mathcal C}_S\to \overline{\mathcal C}_S,
\]
where $\overline{\mathcal C}_S$ may contain a Gorenstein elliptic singularity \cite{SMY1}. Given a family of tropical curves, the function $\delta$ provides a non-negative real number for each point on the base. We may therefore consider the ``\textbf{circle}'' of points on each fiber whose distance from the core is equal to $\delta$. A partial destabilization, or ``bubbling'', introduces $2$-pointed rational curves into the curve by adding vertices to the dual graph at the points that lie on this circle (the crosses at height $\delta$ in Figure \ref{fig:lambdaexample}); we denote this new tropical curve by $\widetilde{\plC}$. The open disc of radius $\delta$, i.e. the locus where $\lambda<\delta$, is then contracted to a Gorenstein singularity in $\overline{\mathcal C}_S$. The number of branches is equal to the number of vertices of $\widetilde{\sqC}$ which lie on the circle. %This data uniquely determines the singularity.

The space of stable maps $\widetilde\VZ_{1,n}(\PP^N,d)$ from the universal nodal curve $\mathcal C$ over the stack $\mathfrak M_{1,n}^{\mathrm{cen}}$ is Deligne-Mumford and proper, with projective coarse moduli space. A compatibility condition between the alignment and the map is required, namely that the circle of radius $\delta$ passes through \emph{at least one component of positive degree}. A stable map in this space is said to satisfy the \textbf{factorisation condition} if the map $\widetilde{\mathcal C}\to \mathcal C\to \mathbb P^N$ factors through the contraction $\widetilde{\mathcal C}\to \overline{\mathcal C}$. Maps that are not constant on any genus $1$ subcurve of $\mathcal C$ automatically satisfy this condition; indeed, the contraction is an isomorphism in this case.

@@

\begin{remark}[Constant maps]\label{rem: constant-maps}
When the degree of the map is $1$, maps from smooth genus $1$ curves to $\PP^N$ are constant. The interior of the moduli space of stable maps in this case is empty, though the compactification by stable maps is non-empty. The space of maps satisfying the factorisation condition is also empty. We do not indulge on this any further. 
\end{remark}

@@

\subsection{Relative geometry: compactification} Fix a hyperplane $H\subseteq \mathbb P^N$. Let $\alpha$ be a partition of the degree $d>0$. Consider the moduli space $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$ of maps from smooth genus $1$ curves $C\to \mathbb P^N$ that meet $H$ at finitely many marked points with vanishing orders given by the partition $\alpha$. This is a smooth, but non-proper Deligne--Mumford stack.

@@

In order to make the obstruction theory more geometric, we expand the target. To elucidate the connection with the static target, consider a logarithmic stable map $[C\to (\mathbb P^N,H)]$ over $\mathrm{Spec} \ (\mathbb N\to \mathbb C)$. At the level of tropicalizations, there is a map of cone complexes:
\[
\plC\to \mathbb R_{\geq 0}.
\]
Choose a subdivision of $\mathbb R_{\geq 0}$ whose vertices are the images of vertices of $\plC$. Pull this subdivision back to $\plC$ by subdividing it along the preimage of the vertices of $\mathbb R_{\geq 0}$. Denote the resulting map $\widetilde \plC \to \widetilde{\mathbb R}_{\geq 0}$. These subdivisions induce logarithmic modifications:
\[
\widetilde C\to \mathbb P^N[s],
\]
see~\cite{AW}. Here the latter is the $s$-times iterated deformation to the normal cone of $H$ in $\mathbb P^N$. The irreducible components of the target are in bijection with the vertices in $\widetilde{\mathbb R}_{\geq 0}$. The curve is modified by adding rational components for the newly introduced vertices.

@@

Globally, Kim constructs a moduli space of logarithmic maps to expanded degenerations, which on logarithmic points, reduces to the above construction~\cite{KimLog}. Kim's space is identified with a subcategory of the Abramovich--Chen--Gross--Siebert space (as stacks over logarithmic schemes), and its dual minimal monoids are cones in a subdivision of the dual minimal cones of the unexpanded space, see~\cite[\S~2]{R19}.

@@

\subsection{Relative geometry: factorisation}\label{subsection factorisation} Let $[C\to \mathbb P^N[s]\to \mathbb P^N]$ be a centrally aligned logarithmic map. Recall that $\mathbb P^N[s]$ consists of a union of $\mathbb P^N$ with $s$ copies of the projective bundle $\mathbb{P} = \mathbb P(\mathcal O_H\oplus \mathcal O_H(1))$. We refer to these latter components as \textbf{the higher levels}. Thus, we will say that a subcurve $D\subseteq C$ \textbf{maps to higher level} if the composite $C\to \mathbb P^N$ maps $D$ into $H\subset \mathbb P^N$.

Let $D_F\subseteq C$ be the maximal genus $1$ subcurve which is mapped to a higher level and contracted by the map $C\to \mathbb P^N[s]$. Let $\delta_F$ denote the distance from the core to the nearest component of $C \setminus D_F$. Similarly let $D_B\subseteq C$ be the maximal genus $1$ subcurve that is contracted by the collapsed map to $\mathbb P^N$, and let $\delta_B$ be the associated radius. This coincides with the radius of the underlying map to $\mathbb P^N$. Of course, $\delta_F\leq \delta_B$, so $(C,\delta_F)$ is centrally aligned as well.

@@

%A Gorenstein elliptic singularity has a number of \textbf{branches}. Given a curve $\overline C$ of genus $1$ with such a singularity, define a branch of $\overline C$ to be an irreducible component containing a branch of the singularity. If the curve $C$ is nodal, then its branches will be understood to be the components of its minimal arithmetic genus $1$ subcurve, i.e. its core.
Recall that we name \emph{core} the minimal genus $1$ subcurve of a Gorenstein curve of arithmetic genus $1$, so the latter decomposes into the nodal union of the core and a (possibly empty) rational forest. If the curve contains an elliptic $m$-fold point, we say a map is non-constant on the core if at least one branch of the singularity is not contracted.
The following condition identifies the locus of smoothable maps.

\begin{definition}
The map $f\colon C\to \mathbb P^N[s]$ \textbf{factors completely} if 
\begin{enumerate}
\item $f$ factors through $\overline C_F$ and it is non-constant on its core,
\item letting $f_B\colon C\to\PP^N$ denote the composite of $f$ with the collapsing map $\mathbb P^N[s]\to\mathbb P^N$, the map $f_B$ factors through $\overline C_B$ and it is non-constant on its core.
\end{enumerate}
\end{definition}

\begin{rem}
 The factorisation conditions are understood to be satisfied when $C\to \mathbb P^N[s]$ (resp. $C\to \mathbb P^N$) does not contract a genus $1$ subcurve. 
\end{rem}

Let
\begin{equation*} \VZ_{1,\alpha}(\mathbb P^N|H,d) \subseteq \widetilde\VZ_{1,\alpha}(\PP^N|H,d) \end{equation*}
be the stack of maps from centrally aligned curves to expansions that factor completely.
